Leaf traits drive differences in biomass partitioning among major plant functional types

### 2015-01-22 manuscript.Rnw
table refs fixed

An advantage of eq. \ref{eq:M1} is that the second term absorbs nearly all of the size related variation in LMF (Tables 1 and S1) and is otherwise fairly conserved between species\citep{chave2005} and across PFTs (Fig. S4). As shown in Fig. 2b and Fig. 2d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Table 1 and S1). $M_F/A_S$ is thus a useful proxy for LMF that can be compared across species and sites.

We analyzed climate effects in two ways: with quantitative variables (mean annual precipitation, MAP; mean annual temperature, MAT), and by biome (tropical / temperate/ boreal), a simple classification taking into account both MAT and MAP (Fig. S1). Regardless of how we analyzed the data, we found only weak climatic effects on biomass partitioning within PFTs (Fig. 3). Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by only 0.01 - 0.06, see Table 2). Likewise, we found weak and inconsistent effects of climate variables (MAP and MAT) (Fig. 3b-c). These variables explained little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by 0.01 and 0.11, respectively, see Table 1).

The finding that climate does not appreciably affect biomass partitioning is good news for models that parameterize the globe by PFT. Yet, these results contrast those of a recent study where LMF was found to increase with MAT across the globe \citep{reich2014}. One explanation for these differences is that our measurements were taken on individual plants, whereas in the other study they were on whole stands. This suggests climatic effects on stand-level biomass partitioning occur primarily by altering the size-distribution and stand density (number of individuals per unit area), rather than partitioning within individual plants. With the caveats that there is still variation between species and studies unaccounted for (Fig. S5), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Tables 1 and 2).

@@

<<results='asis'>>=
print(xtable(table_varpart2, label="varpart2",
             caption="\\textbf{Explained variance by plant functional type (PFT), plant height (H) and biome (B), for five whole-plant variables.} Each variable was added to a linear mixed-effects model, using species within dataset as a random effect. The $R^2$ shown is that explained by the fixed effects only. All fixed effects were highly significant (P \\textless 0.001). In addition to the fixed effects shown, all interactions were added to each of the models."),
      include.rownames=FALSE, caption.placement="top", sanitize.text.function=function(x){x})
@

@@

<<results='asis'>>=
print(xtable(table_gamr2MATMAP, label="table/gamr2table",
      caption="\\textbf{Explained variance by plant functional type (PFT), plant height (H), mean annual temperature (MAT) and mean annual precipitation (MAP), for five whole-plant variables.} Each variable was added as a smooth term in a generalized additive model, with the exception of PFT (a categorical variable)."),
      include.rownames=FALSE, caption.placement="top", sanitize.text.function=function(x){x})
@

>>>>

An advantage of eq. \ref{eq:M1} is that the second term absorbs nearly all of the size related variation in LMF (Tables \ref{table/varpartB} and \ref{table/varpartMAPMAT}) and is otherwise fairly conserved between species\citep{chave2005} and across PFTs (Fig. S4). As shown in Fig. 2b and Fig. 2d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Tables \ref{table/varpartB} and \ref{table/varpartMAPMAT}). $M_F/A_S$ is thus a useful proxy for LMF that can be compared across species and sites.

We analyzed climate effects in two ways: with quantitative variables (mean annual precipitation, MAP; mean annual temperature, MAT), and by biome (tropical / temperate/ boreal), a simple classification taking into account both MAT and MAP (Fig. S1). Regardless of how we analyzed the data, we found only weak climatic effects on biomass partitioning within PFTs (Fig. 3). Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by only 0.01 - 0.06, see Table \ref{table/varpartB}). Likewise, we found weak and inconsistent effects of climate variables (MAP and MAT) (Fig. 3b-c). These variables explained little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by 0.01 and 0.11, respectively, see Table \ref{table/varpartMAPMAT}).

The finding that climate does not appreciably affect biomass partitioning is good news for models that parameterize the globe by PFT. Yet, these results contrast those of a recent study where LMF was found to increase with MAT across the globe \citep{reich2014}. One explanation for these differences is that our measurements were taken on individual plants, whereas in the other study they were on whole stands. This suggests climatic effects on stand-level biomass partitioning occur primarily by altering the size-distribution and stand density (number of individuals per unit area), rather than partitioning within individual plants. With the caveats that there is still variation between species and studies unaccounted for (Fig. S5), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Tables \ref{table/varpartB} and \ref{table/varpartMAPMAT}).

@@

<<results='asis'>>=
print(xtable(table_varpart2, label="table/varpartB",
             caption="\\textbf{Explained variance by plant functional type (PFT), plant height (H) and biome (B), for five whole-plant variables.} Each variable was added to a linear mixed-effects model, using species within dataset as a random effect. The $R^2$ shown is that explained by the fixed effects only. All fixed effects were highly significant (P \\textless 0.001). In addition to the fixed effects shown, all interactions were added to each of the models."),
      include.rownames=FALSE, caption.placement="top", sanitize.text.function=function(x){x})
@

@@

<<results='asis'>>=
print(xtable(table_gamr2MATMAP, label="table/varpartMAPMAT",
      caption="\\textbf{Explained variance by plant functional type (PFT), plant height (H), mean annual temperature (MAT) and mean annual precipitation (MAP), for five whole-plant variables.} Each variable was added as a smooth term in a generalized additive model, with the exception of PFT (a categorical variable)."),
      include.rownames=FALSE, caption.placement="top", sanitize.text.function=function(x){x})
@
